---
title: "model_analysis"
author: "Irsa Ashraf"
date: "2023-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tseries)
library(forecast)
library(TSA)
library(vars)
library(knitr)
```

Read data and check correlations of variables with Robbery 
```{r}
merge_data <- read.csv('merge_data.csv')

vars <- c('ROBBERY', 'avg_temp', 'snowy_days', 'rainy_days', 'unemp')
data <- merge_data[vars]
cor(data)
```
 
Model 1: robb, avg_temp
Model 2: robb, snowy days
Model 3: robb, unemp
Model 4: robb, avg_temp, unemp
Model 5: robb, snowy_days, unemp


TRAIN TEST SPLIT 
```{r}
# CHANGE TRAIN AND TEST
# robbery 
robbery_tsdf <- ts(data$ROBBERY, frequency = 12, start= 2001)

train_robbery <- window(robbery_tsdf, start = c(2001, 1), end = c(2021, 12))
test_robbery <- window(robbery_tsdf, start=c(2022, 1), end=c(2022, 12))

# avg_temp
temp_tsdf <- ts(data$avg_temp, frequency = 12, start= 2001)

train_temp <- window(temp_tsdf, start = c(2001, 1), end = c(2021, 12))
test_temp <- window(temp_tsdf, start=c(2022, 1), end=c(2022, 12))

# Unemployment
unemp_tsdf <- ts(data$unemp, frequency = 12, start= 2001)

train_unemp <- window(unemp_tsdf, start = c(2001, 1), end = c(2021, 12))
test_unemp <- window(unemp_tsdf, start=c(2022, 1), end=c(2022, 12))

# snowy_days
sd_tsdf <- ts(data$snowy_days, frequency = 12, start= 2001)

train_sd <- window(sd_tsdf, start = c(2001, 1), end = c(2021, 12))
test_sd <- window(sd_tsdf, start=c(2022, 1), end=c(2022, 12))

# rainy_days
rd_tsdf <- ts(data$rainy_days, frequency = 12, start= 2001)

train_rd <- window(rd_tsdf, start = c(2001, 1), end = c(2021, 12))
test_rd <- window(rd_tsdf, start=c(2022, 1), end=c(2022, 12))

```


```{r}
# Function for back transformation 

back_transform_fcsts <- function(fcst_data, train_deseas, boxcox_train, train_data) {
  # add trend back 
  xi <- train_deseas[length(train_deseas)]
  diffinv_trend_fcst <- diffinv(fcst_data, differences=1, xi=xi)
  # add seasonality back 
  beg <- length(boxcox_train)-11
  end <- length(boxcox_train)
  diffinv_seas_fcst <- diffinv(diffinv_trend_fcst, lag=12, xi=boxcox_train[beg:end])
  # get rid of boxcox transformation
  fcst_reveresed <- InvBoxCox(diffinv_seas_fcst, lambda = BoxCox.lambda(train_data))
  return(fcst_reveresed)
  }
```


Data Preparation

```{r}
# ROBBERY
# For the train dataset
# BoxCox
lambda_robb = BoxCox.lambda(train_robbery)
robb_boxcox_train <- BoxCox(train_robbery, lambda = lambda_robb)

# differencing 
# kpss test p value = 0.01 so not stationary
print(kpss.test(robb_boxcox_train))

robb_train_deseas <- diff(robb_boxcox_train, lag = 12)
print(kpss.test(robb_train_deseas))
robb_train_stationary <- diff(robb_train_deseas, difference = 1)
tsdisplay(robb_train_stationary)
print(kpss.test(robb_train_stationary))
# p-value = 0.1

# Repeat for the test datatset?

```

A simple model that incorporating lags 1 and 12 is sufficient (session 4, slide 34)
d = 1 and D = 1
q = 1 and Q = 1


AVG_TEMP
BoxCox tranformation 
```{r}
# For the train dataset
# BoxCox
lambda_temp = BoxCox.lambda(train_temp)
temp_boxcox_train <- BoxCox(train_temp, lambda = lambda_temp)

tsdisplay(temp_boxcox_train, ylab="Avg Temperature", xlab="Time")

# differencing 
# kpss test p value = 0.1 so stationary
print(kpss.test(temp_boxcox_train))
print(adf.test(temp_boxcox_train))

temp_train_deseas <- diff(temp_boxcox_train, lag = 12)
print(kpss.test(temp_train_deseas))
temp_train_stationary <- diff(temp_train_deseas, difference = 1)
tsdisplay(temp_train_stationary)
print(kpss.test(temp_train_stationary))
# p-value = 0.1

# Repeat for the test datatset?
```

UNEMPLOYMENT 
```{r}
# For the train dataset
# BoxCox
lambda_unemp = BoxCox.lambda(train_unemp)
unemp_boxcox_train <- BoxCox(train_unemp, lambda = lambda_unemp)

tsdisplay(unemp_boxcox_train, ylab="Unemployment", xlab="Time")

# differencing 
# kpss test p value = 0.1 so stationary
print(kpss.test(unemp_boxcox_train))
print(adf.test(unemp_boxcox_train))

unemp_train_deseas <- diff(unemp_boxcox_train, lag = 12)
print(kpss.test(unemp_train_deseas))
unemp_train_stationary <- diff(unemp_train_deseas, difference = 1)
tsdisplay(unemp_train_stationary)
print(kpss.test(unemp_train_stationary))
# p-value = 0.1

# Repeat for the test datatset?

```

SNOWY DAYS 

```{r}
# For the train dataset
# BoxCox
lambda_sd = BoxCox.lambda(train_sd)
sd_boxcox_train <- BoxCox(train_sd, lambda = lambda_sd)

tsdisplay(sd_boxcox_train, ylab="Snowy Days", xlab="Time")

# differencing 
# kpss test p value = 0.1 so stationary
print(kpss.test(sd_boxcox_train))
print(adf.test(sd_boxcox_train))

sd_train_deseas <- diff(sd_boxcox_train, lag = 12)
print(kpss.test(sd_train_deseas))
sd_train_stationary <- diff(sd_train_deseas, difference = 1)
tsdisplay(sd_train_stationary)
print(kpss.test(sd_train_stationary))
# p-value = 0.1

```

RAINY DAYS 
```{r}
# For the train dataset
# BoxCox
lambda_rd = BoxCox.lambda(train_rd)
rd_boxcox_train <- BoxCox(train_rd, lambda = lambda_rd)

tsdisplay(rd_boxcox_train, ylab="Rainy Days", xlab="Time")

# differencing 
# kpss test p value = 0.1 so stationary
print(kpss.test(rd_boxcox_train))
print(adf.test(rd_boxcox_train))

rd_train_deseas <- diff(rd_boxcox_train, lag = 12)
print(kpss.test(rd_train_deseas))
rd_train_stationary <- diff(rd_train_deseas, difference = 1)
tsdisplay(rd_train_stationary)
print(kpss.test(rd_train_stationary))
# p-value = 0.1

# Repeat for the test datatset?
```

FINAL MODELS 
(presentation order)
MODEL 1: robb, avg_temp
p=5 better than p=4
```{r}
p_ = 5
```


```{r}
# BIND VARIABLES TOGETEHR TO CREATE THE SYSTEM
m1_variables <- cbind(robb_train_stationary, temp_train_stationary)

# USE VARSELECT TO SELECT THE LAG ORDER

# value to use for type?
lagselect_m1 <- VARselect(m1_variables, type = "both")
print(lagselect_m1$selection)

# Apply the model with different values of p

var_4_m1 <- VAR(m1_variables, p =p_, type = "both")
serial.test(var_4_m1, lags.pt=12)
# Used p = _ because of ......
```
```{r}
forecasts_m1 <- forecast(var_4_m1, h=12)

forecasts_m1 %>%
  autoplot(robbery_tsdf) + xlab("Year")

# PLOT FORECASTS 
# ROBBERY

robb_forecasts_m1 <- forecasts_m1$forecast$robb_train_stationary$mean
robb_fcst_reveresed_m1 <- back_transform_fcsts(robb_forecasts_m1, robb_train_deseas, robb_boxcox_train, train_robbery)
model_1_fcst <- window(robb_fcst_reveresed_m1, start=c(2022, 1), end=c(2022, 12))
autoplot(robbery_tsdf) + 
  autolayer(model_1_fcst)

# Caluclate metrics 
print(accuracy(model_1_fcst, test_robbery))

# TEMPERATURE
temp_forecasts_m1 <- forecasts_m1$forecast$temp_train_stationary$mean
temp_fcst_reveresed_m1 <- back_transform_fcsts(temp_forecasts_m1, temp_train_deseas, temp_boxcox_train, train_temp)

autoplot(temp_tsdf) + 
  autolayer(temp_fcst_reveresed_m1)

# Caluclate metrics 
print(accuracy(temp_fcst_reveresed_m1, test_temp))
```

MODEL 2: robb, snowy_days

```{r}
# BIND VARIABLES TOGETEHR TO CREATE THE SYSTEM
m2_variables <- cbind(robb_train_stationary, sd_train_stationary)

# USE VARSELECT TO SELECT THE LAG ORDER

# value to use for type?
lagselect_m2 <- VARselect(m2_variables, type = "both")
print(lagselect_m2$selection)

# Apply the model with different values of p

var_4_m2 <- VAR(m2_variables, p =p_, type = "both")
serial.test(var_4_m2, lags.pt=12)
# Used p = _ because of ......
```
```{r}
forecasts_m2 <- forecast(var_4_m2, h=12)

forecasts_m2 %>%
  autoplot(robbery_tsdf) + xlab("Year")

# FORECAST PLOTS AND METRICS 
# ROBBERY

robb_forecasts_m2 <- forecasts_m2$forecast$robb_train_stationary$mean
robb_fcst_reveresed_m2 <- back_transform_fcsts(robb_forecasts_m2, robb_train_deseas, robb_boxcox_train, train_robbery)

model_2_fcst <- window(robb_fcst_reveresed_m2, start=c(2022, 1), end=c(2022, 12))
autoplot(robbery_tsdf) + 
  autolayer(model_2_fcst)

# Caluclate metrics 
print(accuracy(model_2_fcst, test_robbery))

# SNOWY DAYS
sd_forecasts_m2 <- forecasts_m2$forecast$sd_train_stationary$mean
sd_fcst_reveresed_m2 <- back_transform_fcsts(sd_forecasts_m2, sd_train_deseas, sd_boxcox_train, train_sd)

autoplot(sd_tsdf) + 
  autolayer(sd_fcst_reveresed_m2)

# Caluclate metrics 
print(accuracy(sd_fcst_reveresed_m2, test_sd))
```

MODEL 3: robb, unemp

```{r}
# BIND VARIABLES TOGETEHR TO CREATE THE SYSTEM
m3_variables <- cbind(robb_train_stationary, unemp_train_stationary)

# USE VARSELECT TO SELECT THE LAG ORDER

# value to use for type?
lagselect_m3 <- VARselect(m3_variables, type = "both")
print(lagselect_m3$selection)

# Apply the model with different values of p

var_4_m3 <- VAR(m3_variables, p =p_, type = "both")
serial.test(var_4_m3, lags.pt=12)
# Used p = _ because of ......

# FORECASTS AND PLOTS 
forecasts_m3 <- forecast(var_4_m3, h=12)

forecasts_m3 %>%
  autoplot(robbery_tsdf) + xlab("Year")

# ROBBERY

robb_forecasts_m3 <- forecasts_m3$forecast$robb_train_stationary$mean
robb_fcst_reveresed_m3 <- back_transform_fcsts(robb_forecasts_m3, robb_train_deseas, robb_boxcox_train, train_robbery)

model_3_fcst <- window(robb_fcst_reveresed_m3, start=c(2022, 1), end=c(2022, 12))
autoplot(robbery_tsdf) + 
  autolayer(model_3_fcst)

# Caluclate metrics 
print(accuracy(model_3_fcst, test_robbery))

# UNEMPLOYMENT

unemp_forecasts_m3 <- forecasts_m3$forecast$unemp_train_stationary$mean
unemp_fcst_reveresed_m3 <- back_transform_fcsts(unemp_forecasts_m3, unemp_train_deseas, unemp_boxcox_train, train_unemp)

autoplot(unemp_tsdf) + 
  autolayer(unemp_fcst_reveresed_m3)

# Caluclate metrics 
print(accuracy(unemp_fcst_reveresed_m3, test_unemp))
```


MODEL 4: robb, avg_temp, unemp
```{r}
# BIND VARIABLES TOGETEHR TO CREATE THE SYSTEM
m4_variables <- cbind(robb_train_stationary, temp_train_stationary, unemp_train_stationary)

# USE VARSELECT TO SELECT THE LAG ORDER
lagselect_m4 <- VARselect(m4_variables, type = "both")
print(lagselect_m4$selection)

var_4_m4 <- VAR(m4_variables, p = p_, type = "both")
serial.test(var_4_m4, lags.pt=12)

# FORECAST
forecasts_m4 <- forecast(var_4_m4, h=12)

forecasts_m4 %>%
  autoplot(robbery_tsdf) + xlab("Year")

# PLOT FORECASTS 
# ROBBERY

# BACK TRANSFORM SNOWY DAYS AND CALCULATE RMSE WITH TEST DATASET
robb_forecasts_m4 <- forecasts_m4$forecast$robb_train_stationary$mean
robb_fcst_reveresed_m4 <- back_transform_fcsts(robb_forecasts_m4, robb_train_deseas, robb_boxcox_train, train_robbery)

# forecast <- robb_fcst_reveresed_m4
model_4_fcst <- window(robb_fcst_reveresed_m4, start=c(2022, 1), end=c(2022, 12))
robberies <- window(robbery_tsdf, start=c(2010, 1), end=c(2022, 12))
autoplot(robberies) + 
  autolayer(model_4_fcst)

# Caluclate metrics 
print(accuracy(model_4_fcst, test_robbery))

# TEMPERATURE 
# BACK TRANSFORM SNOWY DAYS AND CALCULATE RMSE WITH TEST DATASET
temp_forecasts_m4 <- forecasts_m4$forecast$temp_train_stationary$mean
temp_fcst_reveresed_m4 <- back_transform_fcsts(temp_forecasts_m4, temp_train_deseas, temp_boxcox_train, train_temp)

temp_model_4_fcst <- window(temp_fcst_reveresed_m4, start=c(2022, 1), end=c(2022, 12))
forecast <- temp_model_4_fcst
temperature <- window(temp_tsdf, start=c(2010, 1), end=c(2022, 12))

autoplot(temperature) + 
  autolayer(forecast)

# Caluclate metrics 
print(accuracy(temp_model_4_fcst, test_temp))

# UNEMPLOYMENT 
unemp_forecasts_m4 <- forecasts_m4$forecast$unemp_train_stationary$mean
unemp_fcst_reveresed_m4 <- back_transform_fcsts(unemp_forecasts_m4, unemp_train_deseas, unemp_boxcox_train, train_unemp)

unemp_model_4_fcst <- window(unemp_fcst_reveresed_m4, start=c(2022, 1), end=c(2022, 12))
forecast <- unemp_model_4_fcst
unemployment <- window(unemp_tsdf, start=c(2010, 1), end=c(2022, 12))
autoplot(unemployment) + 
  autolayer(forecast)

# Caluclate metrics 
print(accuracy(unemp_fcst_reveresed_m4, test_unemp))

```

```{r}
checkresiduals(resid(var_4_m4)[, 1])
```

```{r}
resid_m4 <- resid(var_4_m4)[, 1]
print(Box.test(resid_m4))
```
p-value greater than 0.05 (p=0.9), faill to reject the null hypothesis - no serial correlation


MODEL 5: robb, snowy_days, unemp

```{r}
# Model 5: robb, avg_temp, snowy_days, unemp
# BIND VARIABLES TOGETEHR TO CREATE THE SYSTEM
m5_variables <- cbind(robb_train_stationary, sd_train_stationary, unemp_train_stationary)

# USE VARSELECT TO SELECT THE LAG ORDER

# value to use for type?
lagselect_m5 <- VARselect(m5_variables, type = "both")
print(lagselect_m5$selection)

# Apply the model with different values of p
var_4_m5 <- VAR(m5_variables, p = p_, type = "both")
# serial.test(var_4, lags.pt=12)

# FORECAST
forecasts_m5 <- forecast(var_4_m5, h=12)

forecasts_m5 %>%
  autoplot(robbery_tsdf) + xlab("Year")

# PLOT FORECASTS 
# ROBBERY

# BACK TRANSFORM SNOWY DAYS AND CALCULATE RMSE WITH TEST DATASET
robb_forecasts_m5 <- forecasts_m5$forecast$robb_train_stationary$mean
robb_fcst_reveresed_m5 <- back_transform_fcsts(robb_forecasts_m5, robb_train_deseas, robb_boxcox_train, train_robbery)

model_5_fcst <- window(robb_fcst_reveresed_m5, start=c(2022, 1), end=c(2022, 12))
autoplot(robbery_tsdf) + 
  autolayer(model_5_fcst)

# Caluclate metrics 
print(accuracy(model_5_fcst, test_robbery))


# UNEMPLOYMENT 
unemp_forecasts_m5 <- forecasts_m5$forecast$unemp_train_stationary$mean
unemp_fcst_reveresed_m5 <- back_transform_fcsts(unemp_forecasts_m5, unemp_train_deseas, unemp_boxcox_train, train_unemp)

autoplot(unemp_tsdf) + 
  autolayer(unemp_fcst_reveresed_m5)

# Caluclate metrics 
print(accuracy(unemp_fcst_reveresed_m5, test_unemp))

# SNOWY DAYS 
sd_forecasts_m5 <- forecasts_m5$forecast$sd_train_stationary$mean
sd_fcst_reveresed_m5 <- back_transform_fcsts(sd_forecasts_m5, sd_train_deseas, sd_boxcox_train, train_sd)

autoplot(sd_tsdf) + 
  autolayer(sd_fcst_reveresed_m5)

# Caluclate metrics 
print(accuracy(sd_fcst_reveresed_m5, test_sd))


```
PLOTTING ALL MODEL FORECASTS

```{r}
# robbery_segment <- window(robbery_tsdf, start = c(2015, 1), end=c(2022, 12))

#model_1 <- robb_fcst_reveresed_m1
model_1 <- window(robb_fcst_reveresed_m1, start=c(2022, 1), end=c(2022, 12))
#model_2 <- robb_fcst_reveresed_m2
model_2 <- window(robb_fcst_reveresed_m2, start=c(2022, 1), end=c(2022, 12))
# model_3 <- robb_fcst_reveresed_m3
model_3 <- window(robb_fcst_reveresed_m3, start=c(2022, 1), end=c(2022, 12))
# model_4 <- robb_fcst_reveresed_m4
model_4 <- window(robb_fcst_reveresed_m4, start=c(2022, 1), end=c(2022, 12))
# model_5 <- robb_fcst_reveresed_m5
model_5 <- window(robb_fcst_reveresed_m5, start=c(2022, 1), end=c(2022, 12))
robberies <- window(robbery_tsdf, start=c(2016, 1), end=c(2022, 12))

autoplot(robberies) + 
  autolayer(model_1) +
  autolayer(model_2) +
  autolayer(model_3) + 
  autolayer(model_4) + 
  autolayer(model_5)
```

